CrystalPlan: an Experiment Planning Tool for Crystallography

### 2010-09-21 docs/CrystalPlan_paper/paper.tex
Refs #182: bibliography and figures set up.

Existing software tools [ADD SOME NAMES HERE – see my notes from ACA2010] 
were designed for X-ray experiments where a single incident
wavelength and a large number of sample orientations or sweeps are used to
collect data. These programs cannot be used for TOPAZ, 
since as a time-of-flight instrument, it is able to collect data
from a wide bandwidth of incident neutrons. 
Additionally, this means that relatively few sample orientations are needed in
order to complete a measurement: whereas an X-ray measurement might have
hundreds or thousands of individual frames, a TOPAZ run may have as few as 10.

The goal of CrystalPlan was to make an easy-to-use software tool to aid users in
planning their experiments. By quickly making an experimental plan that will be
sufficient for the needs of the scientist (enough peaks measured), but without
spending time needlessly measuring the same region in reciprocal space,
CrystalPlan will greatly increase the scientific productivity and throughput of
TOPAZ. More users will be able to measure more samples, faster.

@@

CrystalPlan is designed to be able to use a variety of sample orientation
goniometers. A base Goniometer class is subclassed for specific types of
goniometers, allowing a unified API: in a way, it is a “universal goniometer”,
which uses the conventional goniometer angles of $\phi$, $\chi$, $\omega$ as
defined in \cite{knuth84}. Each class inheriting from Goniometer can convert
the desired phi, chi, omega angles to the internal motor positions it requires. The API allows
for limited goniometers: for example, TOPAZ is currently equipped with a
goniometer where chi is fixed at 45 degrees. CrystalPlan handles this by
limiting the available sample orientation angles accessible to the user.

The data analysis program ISAW is used to calculate the the sample mounting
orientation UB matrix, which is saved to a text file along with the lattice
parameters. CrystalPlan can load this UB matrix file and will use that UB matrix
in coverage calculations.

\subsection{Volume coverage}

\subsubsection{Reciprocal Space Representation}

@@

\subsubsection{Calculating Coverage of One Sample Orientation}

@@

When entering the lattice parameters, the crystal's point group symmetry can
also be entered. This information is then used to calculate the increased
measurement redundancy. A list of $n$ $3\times3$ multiplication matrices is
generated based on the point group; any {hkl} vector can be multiplied by these matrices
to find the n equivalent hkl reflections; for example, for orthorombic (mmm)
symmetry, $n=8$ and 8 multiplication matrices are generated, which multiply
{hkl} to give {±h, ±k, ±l}.

@@

In volume coverage mode, a 4D matrix is generated where each voxel has a list of
n corresponding equivalent voxels. This allows CrystalPlan to quickly compute a
volume coverage map that takes into account crystal symmetry: at each voxel, n
original voxels are checked to see if any of them were measured.

@@

\subsubsection{Design}

@@

\subsubsection{Main Program Workflow}
Some screenshots of the GUI are shown in Figure \ref{fig:screenshots}. The main
window consists of a series of tabs, with workflow proceeding from left to right:

@@

\subsubsection{Reciprocal Space Coverage Visualization}
Figure \ref{fig:volume_view}a) shows a screenshot of the reciprocal space
coverage 3D interface. The 3D view can represent q-space in two ways: volume coverage, or single-peak
coverage.

>>>>

Existing diffraction experiment planning software programs (for example,
CrysAlis Pro from Oxford Diffraction; DIFFRAC from Bruker AXS) were designed for
X-ray experiments where a single  incident wavelength and a large number of
sample orientations or sweeps are used to collect data. 
These programs cannot be used for TOPAZ, since as a time-of-flight instrument, 
it is able to collect data from a wide bandwidth of incident neutrons. 
Additionally, this means that relatively few sample orientations are needed in
order to complete a measurement: whereas an X-ray measurement might have
hundreds or thousands of individual frames, a TOPAZ run may have as few as 10.

The goal of CrystalPlan was to make an easy-to-use software tool to aid users in
planning their experiments. By quickly making an experimental plan that will be
sufficient for the needs of the scientist (enough peaks measured), but without
spending time needlessly measuring the same region in reciprocal space,
CrystalPlan will greatly increase the scientific productivity and throughput of
TOPAZ and other beamlines. More users will be able to measure more samples in
less time.

@@

CrystalPlan is designed to be able to use a variety of sample orientation
goniometers. A base Goniometer class is subclassed for specific types of
goniometers, allowing a unified API. This base class acts as a \emph{universal
goniometer},
which uses the conventional goniometer angles of
$\phi$, $\chi$, $\omega$ as defined in \cite{busing67}.
Each class inheriting from Goniometer can convert
the desired phi, chi, omega angles to the internal motor positions it requires. The API allows
for limited goniometers: for example, TOPAZ is currently equipped with a
goniometer where chi is fixed at 45 degrees. CrystalPlan handles this by
limiting the available sample orientation angles accessible to the user.

From an initial data set, the data analysis program ISAW \cite{Tao2006} is used
to calculate the the sample mounting orientation UB matrix, which is saved to a text file along with the lattice
parameters. CrystalPlan can load this UB matrix file and will use that UB matrix
in coverage calculations.

\subsection{Reciprocal Space Representation.}

@@

\subsection{Calculating Coverage of One Sample Orientation}

@@

When entering the lattice parameters, the crystal's point group symmetry can
also be entered. This information is then used to calculate the increased
measurement redundancy. A list of $n$ $3\times3$ multiplication matrices is
generated based on the point group; any $hkl$ vector can be multiplied by
these matrices to find the $n$ equivalent $hkl$ reflections; for example, for
orthorombic (mmm) symmetry, $n=8$ and 8 multiplication matrices are generated, which multiply
$hkl$ to give $\pm h$, $\pm k$, $\pm l$.

@@

In volume coverage mode, a 4D matrix is generated where each voxel has a list
with $n$ entries, each pointing to a corresponding equivalent voxels. This
allows CrystalPlan to quickly compute a volume coverage map that takes into
account crystal symmetry: at each voxel, all $n$ voxels are checked to
see if any of them were measured.

@@

\subsubsection{Design.}

@@

\subsubsection{Main Program Workflow.}
Some screenshots of the GUI are shown in Figure \ref{fig:screenshots}. The main
window consists of a series of tabs, with workflow proceeding from left to right:

@@

\subsubsection{Reciprocal Space Coverage Visualization.}
Figure \ref{fig:volume_view}a) shows a screenshot of the reciprocal space
coverage 3D interface. The 3D view can represent q-space in two ways: volume coverage, or single-peak
coverage.
